---
title: "NYC Leading Causes of Death"
author: "Paula Bailey"
date: "2022-10-14"
output: 
  html_document:
    toc: yes
    toc_depth: 1
    toc_float: yes
    theme: readable
    df_print: "paged"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background

Looking at my profile, one may believe two things about me - 1. I love large cities like Chicago and New York.  2. I'm only interested in health data set.   I haven't been to either city since I was a child; however, both cities have wealth of open source data.  The city of Chicago has their [portal](https://data.cityofchicago.org/).  I just discovered New York City's [NYCOpenData](https://opendata.cityofnewyork.us/). This site also has an API, so I plan to practice accessing APIs in the near future.   As for the second question, it's more likely because I can find more data on the subject of health and crime.  

This data comes from the NYC OpenData website.  It is on the leading causes of death in New York City by sex and ethnicity since 2007.  The cause of death is derived from the NYC death certificate which is issued for every death that occurs in the city".

After I complete this project, I will be completing it again in python.  Currently, I am in school and R is the language of choice.  But I realize, it is important to know python....practice...practice...


# Ask
This is not a formal project, so there is not a detailed business task (or statement).  I will use the data set to explore, complete some summary statistics and visualizations.

I have the following questions.  What was the cause of death for women in NYC?  What about the top 10 causes? Does race change the top cause for women?  Basically, is it different for Black, White, Hispanic, and Asian women?


# Prepare

This csv file is named `New_York_City_leading_Causes_of_Death`.  It contains 1,272 rows and 7 features where each row is a cause of death.  It is contained on the NYC OpenData platform where it is updated annually by Department of Health and Mental Hygiene (DOHMH).  It was last updated on February 8, 2022. The data is from 2007 to 2019.

The csv has the following features:  


|column Name | description | definitions |
|------------|-------------|-------------|
|誰..Year| Year of Death |      year |
|leading_cause| the unique cause of death |      Leading Cause |                      
|Sex | Sex of decedent|  |
|race/ethnicity | Race of decedent |  |                              
|Deaths | Number of People who died due to this cause |  |
|death_rate |Death rate within sex and race category|  |        
|age_adjusted_death_rate |Age adjusted death rate within sex and race|      Age Adjusted Death Rate|     

Based on the information from the website all of the variables are in plain text.  So, we will need to verify the types of each column after we import the data.

The data and metadata can be found [here.](https://data.cityofnewyork.us/Health/New-York-City-Leading-Causes-of-Death/jb7j-dtam)

After downloading the file, I renamed it to preserve the original file.  During the data cleaning (process), any changes to the data will be saved under a file named ChangeLog.doc   


# Process

```{r library, warning=FALSE, message=FALSE}
library(tidyverse)
```


First, we will read in the csv file.
```{r}
nycCauses <- read.csv("New_York_City_Leading_Causes_of_Death_pmb.csv", na.strings=c("",".","NA"))
```


```{r}
head(nycCauses)
```

Next, the structure of the file.  
```{r}
str(nycCauses)
```

As mentioned above, everything is a character format except for year, including deaths, death.rate, and the adjusted death rate.

I think we should make the following data changes:

  * Leading Cause, Sex, and Race.Ethnicity from CHR to FACTOR
  * Leading Cause, Sex, and Race.Ethnicity from CHR to FACTOR
  * death rates and age adjusted deaths rates from CHR to DOUBLE
  * Change 誰..Year to year (column name)

After our data type changes, our data sets will contain 3 continuous features, 1 discrete feature and 3 categorical features.

Rename column 誰..Year
```{r}
nycCauses <- nycCauses %>% rename(year=誰..Year)
```

Before changing data types, let's look for any NAs.

During this process, I ran into `Warning: NAs introduced by coercionWarning` which indicates the data set has missing information (empty not NA). Hmm, that is interesting.  I was not expecting to see NAs. I resolved this issue by adding `na.strings=c("",".","NA")` during import.

```{r}
colSums(is.na(nycCauses))
```

After reviewing the data set, we can see which descendants have NAs.

  *  Descendants with  "Other" as race do not have Death rate and age adjusted death rate.
  *  The remaining 138 descendants with missing deaths are a combination of  with "other Race" and "Not Stated" as race.
  
Based on the above, I will create a data set descendants with missing deaths removed.  We will name that data frame as `nycCauses_Deaths`.  I will try to complete most of the analysis with that data set. Recall the original data set is 1,272 observations and now we have 1,134.  The difference is 138.

```{r}
nycCauses_Deaths <- nycCauses %>% drop_na(Deaths)
```

Change varaibles Leading causes, sex, and race to factors. Change deaths, death,rate and age adjusted death rate to integer and double, respectively.
```{r}

# update to factors
nycCauses_Deaths$Leading.Cause = factor(nycCauses_Deaths$Leading.Cause)
nycCauses_Deaths$Sex = factor(nycCauses_Deaths$Sex)
nycCauses_Deaths$Race.Ethnicity = factor(nycCauses_Deaths$Race.Ethnicity)

# update to integer and double
nycCauses_Deaths$Deaths = as.integer(nycCauses_Deaths$Deaths)
nycCauses_Deaths$Death.Rate = as.double(nycCauses_Deaths$Death.Rate)
nycCauses_Deaths$Age.Adjusted.Death.Rate = as.double(nycCauses_Deaths$Age.Adjusted.Death.Rate)

```

We will also create a second data set named`nycCause_noNA`, where we remove all NAs which will have 819 observations.

```{r}
nycCauses_noNA   <- nycCauses_Deaths %>% drop_na(Death.Rate, Age.Adjusted.Death.Rate)
```

Let's confirm our changes
```{r}
str(nycCauses_Deaths)
```


```{r}
nycCauses_Deaths %>% select(Deaths, Death.Rate, Age.Adjusted.Death.Rate) %>% summary()
```


```{r}
nycCauses_noNA %>% select(Deaths, Death.Rate, Age.Adjusted.Death.Rate) %>% summary()
```

Let's look at the distinct values for our character values - Leading_Cause, Sex and Race.Ethnicity. I noticed in the str() above for sex, there are four factors listed.  I want to make sure those factors are unique. 
```{r}
n_distinct(nycCauses_Deaths$year)
n_distinct(nycCauses_Deaths$Sex)
n_distinct(nycCauses_Deaths$Leading.Cause)
n_distinct(nycCauses_Deaths$Race.Ethnicity)
```

```{r}
unique(nycCauses_Deaths[["Sex"]])
```

Wow...As we can see,it's a good thing that we looked.  We have Male, M, Female and F. So, we will change the factor of F and M to Female and Male. Since we already created nycCauses_noNA, we will update that files as well.

```{r}
levels(nycCauses_Deaths$Sex)[levels(nycCauses_Deaths$Sex)=='M'] <- 'Male'
levels(nycCauses_Deaths$Sex)[levels(nycCauses_Deaths$Sex)=='F'] <- 'Female'

levels(nycCauses_noNA$Sex)[levels(nycCauses_noNA$Sex)=='M'] <- 'Male'
levels(nycCauses_noNA$Sex)[levels(nycCauses_noNA$Sex)=='F'] <- 'Female'
```

```{r}
unique(nycCauses_Deaths[["year"]])
```

We notice that we are missing data from 2015 - 2018. I'm not sure why we are missing data.  The data dictionary in the `Cause_of_Death_121412.csv` file states years from 2007 - 2016 and the title gives the impression we will have data from 2007.  We have an discrepancy, but we will carry on.


Look at factors for ethnicity.
```{r}
unique(nycCauses_Deaths[["Race.Ethnicity"]])
```
As a best practice, it is good to check all levels of features with factors.  Above we have both Black Non-Hispanic and Non-Hispanic Black and White Non-Hispanic and Non-Hispanic White,  We need to address this so it will not effect our work.

```{r}
levels(nycCauses_Deaths$Race.Ethnicity)[levels(nycCauses_Deaths$Race.Ethnicity)=='Non-Hispanic Black'] <- 'Black Non-Hispanic'
levels(nycCauses_Deaths$Race.Ethnicity)[levels(nycCauses_Deaths$Race.Ethnicity)=='Non-Hispanic White'] <- 'White Non-Hispanic'


levels(nycCauses_noNA$Race.Ethnicity)[levels(nycCauses_noNA$Race.Ethnicity)=='Non-Hispanic Black'] <- 'Black Non-Hispanic'
levels(nycCauses_noNA$Race.Ethnicity)[levels(nycCauses_noNA$Race.Ethnicity)=='Non-Hispanic White'] <- 'White Non-Hispanic'
```


Check to confirm our changes

```{r}
unique(nycCauses_Deaths[["Sex"]])
```

```{r}
unique(nycCauses_Deaths[["Race.Ethnicity"]])
```

Now our data sets seem to be clean.  As mentioned above we will use nycCauses_Deaths and nycCauses_noNA. We will mainly use nycCauses_Deaths, because we were able to maintain the most data.


# Analyze

Question 1 - For black women who died in 2019, how many women died from heart disease or diabetes?
```{r message=FALSE}
nycCauses_Deaths %>% 
  group_by(year, Leading.Cause, Sex, Race.Ethnicity) %>% 
  summarise(mean = mean(Deaths)) %>%
  filter(year == 2019 & Sex == "Female" & grepl('Heart|Diabetes', Leading.Cause) 
         & Race.Ethnicity =='Black Non-Hispanic')
  
```
In 2019, on average 383 black women died from diabetes and 2,483 died from heart disease.


Question 2 - In 2019, what is the top causes of death for women?
```{r message=FALSE}
  nycCauses_Deaths %>% 
        group_by(year, Leading.Cause, Sex) %>% 
        summarise(average = mean(Deaths)) %>%
        filter(year == 2019 & Sex == "Female") %>%
        arrange(desc(average)) %>%
        select(Leading.Cause, average) %>%
        head(10)

```
The top cause of death for women is again heart disease with an average of 1,266 deaths. 


Question 3 - Question two leads us to ask, Is heart disease the leading cause of death for all race/ethnicity in NYC in 2019?  Is it a leading cause by race? We will look at Black, White, Hispanic, Asian and Pacific Islander and Other Race.  

```{r}
nycCauses_Deaths %>% 
      group_by(year, Race.Ethnicity,Leading.Cause) %>% 
      filter(year == 2019 & Sex == "Female" & Race.Ethnicity == 'Black Non-Hispanic') %>%
      summarise(average = mean(Deaths)) %>%
      arrange(desc(average)) %>%
      select(Race.Ethnicity, Leading.Cause, average) 
```
The top cause of death for Black females in 2019 was Diseases of Heart (I00-I09, I11, I13, I20-I51) with an average of 2,483 women dying.  The second highest is Malignant Neoplasms (Cancer: C00-C97)  with 1,753 cases. 


```{r}
nycCauses_Deaths %>% 
      group_by(year, Race.Ethnicity,Leading.Cause) %>% 
      filter(year == 2019 & Sex == "Female" & Race.Ethnicity == 'White Non-Hispanic') %>%
      summarise(average = mean(Deaths)) %>%
      arrange(desc(average)) %>%
      select(Race.Ethnicity, Leading.Cause, average) 
```
For Non-Hispanic White women, the number one cause is Diseases of Heart with average 4,001 deaths. This is followed by Cancer and Chronic Lower Respiratory Diseases. 

```{r}
nycCauses_Deaths %>% 
      group_by(year, Race.Ethnicity,Leading.Cause) %>% 
      filter(year == 2019 & Sex == "Female" & Race.Ethnicity == 'Hispanic') %>%
      summarise(average = mean(Deaths)) %>%
      arrange(desc(average)) %>%
      select(Race.Ethnicity, Leading.Cause, average) 
```
For Hispanic women, the number one cause of death is Diseases of Heart with average 1,574 deaths. This is followed by Cancer and Cerebrovascular Disease. 

```{r}
nycCauses_Deaths %>% 
      group_by(year, Race.Ethnicity,Leading.Cause) %>% 
      filter(year == 2019 & Sex == "Female" & Race.Ethnicity == 'Asian and Pacific Islander') %>%
      summarise(average = mean(Deaths)) %>%
      arrange(desc(average)) %>%
      select(Race.Ethnicity, Leading.Cause, average) 
```
For Asian and Pacific Islander women, the number one cause of death is Diseases of Heart with average 623 deaths. This is followed by Cancer and Cerebrovascular Disease. 

```{r}
nycCauses_Deaths %>% 
      group_by(year, Race.Ethnicity,Leading.Cause) %>% 
      filter(year == 2019 & Sex == "Female" & Race.Ethnicity == 'Other Race/ Ethnicity') %>%
      summarise(average = mean(Deaths)) %>%
      arrange(desc(average)) %>%
      select(Race.Ethnicity, Leading.Cause, average) 
```
For women who identified as other race, Diseases of Heart is also the top cause of death with average 623 deaths. This is followed by Cancer and Diabetes Mellitus. 


# Share

This plot shows that heart disease has been trending down since 2007 as the cause of death for women. However, it still seems to +be the number one cause of death for women.
```{r}
nycCauses_Deaths %>% 
        filter(Sex == "Female" & Leading.Cause == 'Diseases of Heart (I00-I09, I11, I13, I20-I51)') %>%
  
       ggplot( aes(x=year, y = Deaths)) + 
        theme(axis.text.x=element_text(angle=50, hjust=1)) +
        geom_bar(stat = "identity")

```

I have added images of two additional plots that are contained in a file named `Tableau_images.doc.`  Those plots show heart disease has been declining since 2007, but continues to be the number one cause of death for men and women.


# Conclusion

As mentioned, I plan to begin the project using python; however this time I will look at the causes of death for men.  I will also be using the data to complete plots (charts) in Tableau.  Again, another way to practice.

In the meantime, we can see that heart disease has been the leading cause of death for women in the NYC area.  This is regardless their ethnicity.