---
title: "NYC Leading Causes of Death"
author: "Paula Bailey"
date: "2022-10-14"
output: 
  html_document:
    toc: yes
    toc_depth: 1
    toc_float: yes
    theme: readable
    df_print: "paged"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background

Looking at my profile, one may believe two things about me - 1. I love large cities like Chicago and New York 2. I'm only interested in health data set.   I haven't been to either city since I was a child; however, both have wealth of open source data.  The city of Chicago has their [portal](https://data.cityofchicago.org/).  I just discovered New York City's [NYCOpenData](https://opendata.cityofnewyork.us/). This site also has an API, so I plan to practice in the near future.  As for the second question, it's more likely because I can find more data on the subject of health and crime.  

This data comes from the NYC OpenData website.  It is on the leading causes of death in New York City by sex and ethnicity since 2007.  The cause of death is derived from the NYC death certificate which is issued for every death that occurs in the city"

After I complete this project, I will be completing it again in python.  Currently, I am in school and R is the language of choice.  But I realize, it is important to know python. I've completed a program on the DataCamp platform...practice...practice...


# Ask
This is not a formal project, so there is not a detailed business task (or statement).  I will use the data set to explore, complete some summary statistics and visualizations.


# Prepare

This csv file is named `New_York_City_leading_Causes_of_Death`.  It's in a long format with a shape of 1,272 x 7 where each row is a cause of death.  It is contained on the NYC OpenData platform where it is updated annually by Department of Health and Mental Hygiene (DOHMH).  It was last updated on February 8, 2022.

The csv has the following columns (variables):  
Year                     Year of Death                                    year
Leading Cause            the unique cause of death                        leading_cause
Sex                      Sex of decedent                                  sex
Race                     Race of decedent                                 race/ethnicity
Deaths                   Number of People who died due to this cause      deaths
Death Rate               Death rate within sex and race category          death_rate
Age Adjusted Death Rate  Age adjusted death rate within sex and race      age_adjusted_death_rate

Based on the information from the website all of the variables are in plain text.  So, we will need to verify the types of each column.

The data and metadata can be found [here.](https://data.cityofnewyork.us/Health/New-York-City-Leading-Causes-of-Death/jb7j-dtam)

After downloading the file, I renamed it to preserve the original file.  During the data cleaning (process), any changes to the data will be saved under a file named ChangeLog.doc   

# Process

```{r library, warning=FALSE}
library(tidyverse)
```


First, we will read in the csv file.
```{r}
nycCauses <- read.csv("New_York_City_Leading_Causes_of_Death_pmb.csv")
```

Next, the structure of the file.  
```{r}
str(nycCauses)
```

As mentioned above, everything is a character format except for year including deaths, death.rate, and the adjusted death rate.

I think we should make the following data changes:

  * Year changes INT to DATE
  * Leading Cause, Sex, and Race.Ethnicity from CHR to FACTOR
  * deaths from CHR to INT
  * death rates and age adjusted deaths rates from CHR to DOUBLE
  
